Working with the Poisson distribution

### 2021-08-31 text/21/poisson-inferences.Rmd
Added link to poisson-inference.md
Minor fix to poisson-inferences.Rmd
Linked in some references for poisson-inference.md

@@

If Y has a binomial distribution with n trials and probability of success p (X ~ Bin(n,p)), then for large n and small p, you can approximate Y with X ~ Pois($\lambda$=np).

@@

and a crude 95% confidence interval for $\lambda$ is

$X \pm 2 \sqrt{X}$

@@

x ~ Pois($\lambda$T)

where T is known, but $lambda$ is unknown, then a 95% confidence interval for $lambda is

$\frac{X}{T} \pm 2 \frac{\sqrt{X}}{T}$

@@

X ~ Pois(T$\lambda$) and

Y=ln(X), then

E[Y] $\approx$ ln($\lambda T$) and

V[Y] $\approx \frac{1}{\lambda T}$

An approximate 95% confidence interval for ln($\lambda  T$) is

ln(X) $\pm 2 \sqrt{\frac{1}{X}}$

@@

@@

http://www.pmean.com/07/ConfidenceIntervalForRate.html

https://www.health.pa.gov/topics/HealthStatistics/Statistical-Resources/UnderstandingHealthStats/Documents/Confidence_Intervals_for_a_Crude_Rate.pdf

>>>>

If you have series of Poisson random variables,

$X_1, X_2,...,X_n$ all indpendent and all with the same Pois($\lambda$) distribution, then

$\Sigma_{i=1}^n X_i$ ~ Pois($k \lambda$)

These properties and more are described on the [Wikipedia page on the Poisson distribution][pois1].

@@

If Y has a binomial distribution with n trials and probability of success p (X ~ Bin(n,p)), then for large n and small p, you can approximate Y with X ~ Pois($\lambda$=np). Wikipedia has a [nice explanation][pois2] of this.

@@

and a crude $1-\alpha$ confidence interval for $\lambda$ is

$X \pm z_{\alpha/2} \sqrt{X}$

@@

x ~ Pois($\lambda T$)

where T is known, but $\lambda$ is unknown, then a $1-\alpha$ confidence interval for $\lambda$ is

$\frac{X}{T} \pm z_{\alpha/2} \frac{\sqrt{X}}{T}$

I have a [page on my website][simo1] that describes these calculations.

@@

X ~ Pois($\lambda T$) and

$Y = g(X) = ln(X)$, then

$g'(X)=\frac{1}{X}$, and

V[Y] $\approx (g'(\mu))^2 \sigma^2_x = (\frac{1}{\lambda T})^2 \lambda T = \frac{1}{\lambda T}$

An approximate $1-\alpha$ confidence interval for ln($\lambda  T$) is

$ln(X) \pm z_{\alpha/2} \sqrt{\frac{1}{X}}$

@@

The mathematical approach used here is known as the delta method and is [documented in general terms] on Wikipedia.

@@

The Pennsylvania Department of Health has a [web page][penn1] oulining calculations for rates with three simple examples. The infant mortality rate in Butler County in 1984 was 17 deaths out of 1,989 births. This is actually a proportion, but let's treat it as a rate here.

```{r}
x <- 17
n <- 1989
rate <- x/n
rate
```

This rate is difficult to read because all of the leading zeros. So it is common practice to quote it not as a rate per live births, but a rate per 1,000 live births.

```{r}
rate_per_thousand <- rate*1000
rate_per_thousand
```

To construct a confidence interval for the rate, first start with a confidence interval for the count rather than the rate.

```{r}
lo <- 
hi <- x+1.96*sqrt(x)
count_ci <- c(x-1.96*sqrt(x), x+1.96*sqrt(x))
count_ci
```

Then convert from a count to a rate.

```{r}
rate_ci <- count_ci * 1000 / n
rate_ci
```

It's always a good idea to round your results.

```{r}
round(rate_ci, 1)
```

Now let's try it using the improved formula. First build a confidence interval for the log count.

```{r}
log_count <- log(x)
log_ci <- c(log_count-1.96*sqrt(1/x), log_count+1.96*sqrt(1/x))
log_ci
```

Back transform to the original scale.

```{r}
back_transformed_ci <- exp(log_ci)
back_transformed_ci
```

Convert from a count to a rate

```{r}
rate_ci <- back_transformed_ci * 1000 / n
rate_ci
```

And again remember to round your final result.

```{r}
round(rate_ci, 1)
```

Notice that the rate (`r round(rate_per_thousand, 2)`) is not in the middle of the interval. It is slightly closer to the lower limit than the upper limit. This reflects the skewness in the Poisson distribution and provides a slightly more accurate interval.

### On your own

The reference also includes an infant mortality rate for the city of Philadelphia. There were 388 infant deaths out of 24,979 live births. Show that the 95% confidence interval for the infant mortality rate in Philadelphia is 14.0 to 17.1.

Notice, by the way, how much narrower the confidence interval is when it is based on 388 deaths rather than 17.

### References

[simo1]: http://www.pmean.com/07/ConfidenceIntervalForRate.html

[penn1]: https://www.health.pa.gov/topics/HealthStatistics/Statistical-Resources/UnderstandingHealthStats/Documents/Confidence_Intervals_for_a_Crude_Rate.pdf

[delt1]: https://en.wikipedia.org/wiki/Delta_method

[delt2]: https://en.wikipedia.org/wiki/Taylor_expansions_for_the_moments_of_functions_of_random_variables

[pois1]: https://en.wikipedia.org/wiki/Poisson_distribution

[pois2]: https://en.wikipedia.org/wiki/Binomial_distribution#Poisson_approximation

http://www.pmean.com/definitions/or.htm

http://www.pmean.com/01/oddsratio.html

http://www.pmean.com/04/Rates.html

https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005510
